---
title: "MB5370 Module 3 Workshop 3"
author: "Caitlin Buckle"
format: html
editor: visual
---

In this workshop we will assemble a mock metagenome consisting of five bacterial species. This metagenome has been designed to capture some of the key challenges in metagenomic assembly, including variable coverage (some species abundant, others rare - assembling rare taxa is more difficult) and similar sequences (this community contains two vibrio species - similar enough that they are difficult to separate in an assembly)

Created new directory (raw_data) and a symbolic link within the directory to the raw reads file in this directory 
```bash
ln -s /pvol/data/metagenome_assembly/reads_R10_0.5.fastq.gz raw_reads.fastq.gz
```

Starting point - examine the read length distribution in our input data 

```bash
zcat raw_data/raw_reads.fastq.gz | head -n 40000 | bioawk -c fastx '{print $name,length($seq)}' > read_lengths.txt
```
Does say that there is a broken pipe - AI says this is not a problem 

Create an R code chunk to visualise the read lengths 

```{r}
library(tidyverse)

rl_data <- read_tsv("read_lengths.txt",col_names = c("id","length"), show_col_types = FALSE)

ggplot(rl_data, aes(x=length/1000)) + 
  geom_histogram(binwidth = 2) +
  xlab("Read Length Kb") + ylab("Number of Reads")
```

Plot is right skewed, with most reads between 1 and 50kb in length - now going to summarise with N50 statistic (most often used to summarise genome assemblies but can also be used for summarising read length distributions). Useful as unaffected by having large numbers of tiny reads and instead places emphasis on the longer reads in the data. 

```{r}
total_len <- sum(rl_data$length) #creating new object 'total_len' and defining as the sum of the lengths within rl_data 

rl_data %>% 
  arrange(desc(length)) %>% #arrange the reads from longest to shortest
  mutate(cumsum = cumsum(length)) %>% #create a new column with cumulative sum (cumsum) of read lengths (starting from longest as described above)
  filter(cumsum>total_len/2) %>% #filtering the reads to find where the cumulative sum (cumsum) surpasses 50% of the total read volume (total_len) - this is the N50!
  head(n=1)

#N50 of individual reads = 22kb (22322bp length)
```

N50 of individual reads is ~22kb - we expect the N50 of our assembly to be considerably greater than this 

Using flye as assembler - created shell file (run_flye.sh) in flye directory, and submitted job (batch job 386)

While flye is running - Tabulating chromosome lengths and their corresponding GC content using the bioawk program created in gc.awk (txt file in marine-genomics directory)

Use cat to concatenate all the files together and pipe to send their contents to our bioawk program 

```bash
cat /pvol/data/mg/refs/*.fna | bioawk -c fastx -f gc.awk
```

Output: 
CP003591.1      4681111 0.584631
CP045504.1      4388462 0.636383
CP045505.1      51725   0.578966
AP024165.1      3168162 0.44373
AP024166.1      1553697 0.438144
AP024167.1      320083  0.418301
AP024168.1      221076  0.434733
CP098033.1      1819821 0.445875
CP098034.1      3294375 0.447314
CP098035.1      36224   0.485231
NZ_CP053351.1   4560973 0.34633

gc.awk output - name, length, gc content 

slurm-386.out finished, new folder in flye called 'mock' with all the required files inside 

Flye reported outputs: 
    Total length:   18270553
        Fragments:      40
        Fragments N50:  1866219 (~186kb) - IMPORTANT: much larger than our individual reads N50 
        Largest frg:    4681059
        Scaffolds:      0
        Mean coverage:  12

From assembly_info.txt - largest config is config_27 (4681059bp, coverage 21x, also circular)
Circular contigs (from assembly_info.txt) - config_27 (4681059bp), contig_39	(1553714bp), contig_41	(320168bp), contig_40	(221077bp), contig_42	(5088bp) 

Previously used gc.awk to print lengths and GC content for our reference sequences - now do they same for our assembled contigs 

Use gc.awk script to print GC and lengths for the assembly 

```bash
cat ../flye/mock/assembly.fasta | bioawk -c fastx -f ../gc.awk 
```
Run again and send to 'assembly.gc.tsv' file 

```bash
cat ../flye/mock/assembly.fasta | bioawk -c fastx -f ../gc.awk > assembly.gc.tsv
```

GC content is in assembly.gc.tsv and coverage depth in assembly_info.txt - read those and join them together using R 

```{r}
gc <- read_tsv("assembly_check/assembly.gc.tsv",col_names = c("#seq_name","length","GC"))
info <- read_tsv("flye/mock/assembly_info.txt") 
```

Both datasets have a #seq_name column so we can join them together using code below 

```{r}
gc_info <- info %>% 
  left_join(gc)
```

Create a histogram by GC content 

```{r}
gc_info %>% 
  ggplot(aes(x=GC)) + geom_histogram() + ylab("Number of contigs")
```

Histogram by GC content shows several peaks - likely corresponds to major groupings of taxonomically distinct lineages within the data 

Create plot of GC vs coverage helps to identify clusters - first filter data to remove contigs with unrealistically high GC content and very high coverage depth (probs repeats)

```{r}
gc_info %>% 
  filter(GC<0.8) %>% 
  filter(cov.<25) %>% 
  ggplot(aes(x=GC,y=cov.)) + geom_point(aes(size=length)) + ylab("Coverage depth")
```
Contig lengths shown through point size, coverage and GC content on x and y axis - Vibrio alfacsensis contigs cluster together etc (look at the table at the top of the workshop manual for expected GC content for each bacteria in our metagenome)

Addings labels - had to install ggrepel package :) only contigs with more than 25x coverage and less than 8-% GC coverage are shown 
```{r}
library(ggrepel)
gc_info %>% 
  filter(GC<0.8) %>% 
  filter(cov.<25) %>% 
  ggplot(aes(x=GC,y=cov.)) + geom_point(aes(size=length)) + geom_label_repel(aes(label=`#seq_name`))
```

Tools to bin broadly fall into two categories - read binning tools which group the raw (unassembled) reads, or contig binning tools which group assembled contigs. Both cases produce 'bins', which will cntain multiple distinct sequences (reads or contigs) that come the same organism. 

Binning is a very difficult problem and results from binning often contain errors such as placing two sequences together when they come from different taxa (leads to 'contamination') or failing to group sequences that come from the same taxa (lead to fragmentation)

We will perform contig binnings 

New directory (metabat), new input file inside 'depths.txt' 

```bash
cat ../flye/mock/assembly_info.txt  | grep -v '#' | awk '{OFS="\t";print $1,$3}' > depths.txt
```

Create a new version of the assembly file in which the contigs have the same order as those in the depths.txt file - using samtools faidx (allows us to extract entries from a fasta file in whatever order we specify)

```bash
cat depths.txt | awk '{print $1}' | xargs samtools faidx ../flye/mock/assembly.fasta > assembly_ordered.fasta
```

Run metabat 

```bash
metabat2 -i assembly_ordered.fasta -a depths.txt --cvExt -o bins/metabat --seed 5
```

Bins directory will contain multi-fasta files with each of the bins inferred by metabat - quickly inspect using grep command 

```bash
grep '>' bins/*
```
Grep output - 5 bins 
bins/metabat.1.fa:>contig_24
bins/metabat.1.fa:>contig_25
bins/metabat.1.fa:>contig_35
bins/metabat.2.fa:>contig_6
bins/metabat.2.fa:>contig_10
bins/metabat.2.fa:>contig_14
bins/metabat.2.fa:>contig_19
bins/metabat.2.fa:>contig_32
bins/metabat.2.fa:>contig_15
bins/metabat.3.fa:>contig_16
bins/metabat.3.fa:>contig_9
bins/metabat.3.fa:>contig_17
bins/metabat.3.fa:>contig_7
bins/metabat.3.fa:>contig_26
bins/metabat.3.fa:>contig_1
bins/metabat.3.fa:>contig_18
bins/metabat.3.fa:>contig_29
bins/metabat.3.fa:>contig_28
bins/metabat.3.fa:>contig_38
bins/metabat.3.fa:>contig_3
bins/metabat.3.fa:>contig_5
bins/metabat.4.fa:>contig_20
bins/metabat.4.fa:>contig_21
bins/metabat.4.fa:>contig_31
bins/metabat.4.fa:>contig_4
bins/metabat.4.fa:>contig_23
bins/metabat.4.fa:>contig_33
bins/metabat.4.fa:>contig_13
bins/metabat.4.fa:>contig_22
bins/metabat.5.fa:>contig_2
bins/metabat.5.fa:>contig_39
bins/metabat.5.fa:>contig_41
bins/metabat.5.fa:>contig_40

Use checkM to assess quality of our bins 

In new directory (checkm, within Marine-Genomics), copy over bins from metabat

```bash
cp ../metabat/bins/metabat.*.fa bins/
```

Metabat failed to include our longest contig in bins - need to extract this contig from our assembly and include it with out metabat bins 

```bash
samtools faidx ../flye/mock/assembly.fasta contig_39 > bins/contig_39.fa
```

Run the checkm workflow - create new shell file (run_checkm.sh) and submit with sbatch - batch job 404 

```sbatch
sbatch run_checkm.sh
```

checkm_results.txt results 
Bin Id               Marker lineage            # genomes   # markers   # marker sets    0     1    2    3   4   5+   Completeness   Contamination   Strain heterogeneity  
----------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  metabat.5      f__Vibrionaceae (UID4865)           80         922           367         4    899   17   2   0   0       99.93            1.33              26.09          
  metabat.4   c__Deltaproteobacteria (UID3217)       62         280           168         33   247   0    0   0   0       85.86            0.00               0.00          
  metabat.3      f__Vibrionaceae (UID4865)           80         922           367        740   182   0    0   0   0       18.18            0.00               0.00          
  metabat.2             root (UID1)                 5656         56            24         53    3    0    0   0   0       12.50            0.00               0.00          
  metabat.1      f__Vibrionaceae (UID4865)           80         922           367        846    76   0    0   0   0        7.48            0.00               0.00          
  contig_39             root (UID1)                 5656         56            24         56    0    0    0   0   0        0.00            0.00               0.00          
----------------------------------------------------------------------------------------------------------------------------------------------------------------------------

Bin 5 has the highest level of contamination 

Starting to use dataset for Black Band Disease (BBD) - go back into raw_data directory, then make symbolic link to the bbd read data 

```bash
ln -s /pvol/data/metagenome_assembly/bbd_0.1.fastq.gz .
```

Make copy of run_flye.sh file 

```bash 
cp run_flye.sh run_flye_bbd.sh
```

Submit job using sbatch - batch job 412 - this will take a while to run, best to check its working then leave overnight to run 

```bash 
sbatch run_flye_bbd.sh
```




